\section{The ARCTO Library}
\label{sec:arcto-library}

\arcto{} is a GPU compression library for AMD GPUs designed to play
the role that \nvcomp{}~\cite{nvcomp} plays on the NVIDIA side: a
stable, batched, multi-algorithm public C API that scientific
applications can link against in production. The design choices are
made explicit in this section so that the two optimization
contributions presented later (Sections~\ref{sec:optim-chunk-size}
and~\ref{sec:optim-adaptive-tiled}) sit on a defensible infrastructure.

\subsection{Design constraints}
\label{sec:arcto-constraints}

Three constraints shape every \arcto{} decision. First, \emph{the
contribution sits in the optimization layer, not in the algorithmic
cores.} The four compression kernels are inherited from established
references (LZ4, Snappy, and Cascaded from the \nvcomp{}~2.2
codebase; ZFP from its canonical upstream) and treated as fixed black
boxes from \arcto{}'s point of view. This separation reserves the
engineering effort and the experimental contribution of the present
work for the profile-driven layer that surrounds the kernels,
presented in Sections~\ref{sec:optim-chunk-size}
and~\ref{sec:optim-adaptive-tiled}, on top of cores whose correctness
has already been validated by years of CUDA-side production use.
Second, \emph{binary
compatibility and language reach matter}: production scientific
codes routinely link from Fortran and C call sites, and the
target application of the present study, the Fletcher seismic
forward-modelling pipeline~\cite{fletcher2009tti}, is no exception.
Third, \emph{portability across the AMD lineup is mandatory but
must not preclude per-architecture tuning}: the same source tree has
to compile and run on GCN, CDNA~$\{$1,2,3$\}$ and RDNA~3 with
architecture-aware constants, not with runtime arch dispatch that
adds branches inside hot kernels.

\subsection{Porting approach}
\label{sec:arcto-porting}

\arcto{} bootstraps from the open-source nvCOMP~2.2
codebase~\cite{nvcomp}, which provides a batched compression API
that processes multiple independent data chunks in parallel, a
design well-suited to GPU architectures. The codebase contains
well-structured, GPU-optimized implementations of the three target
byte-level algorithms (LZ4, Snappy, Cascaded), and its batched API
model maps directly to the HIP programming model. The porting
methodology follows the standard CUDA-to-HIP translation workflow
documented in the HIP programming guide~\cite{rocm} and proceeds in
three phases.

\emph{API translation.} \arcto{} uses AMD's
\texttt{hipify-perl} tool~\cite{rocm} to perform the initial
mechanical translation of the CUDA source code to HIP. The tool
handles approximately 90\,\% of the API surface automatically; the
remaining manual modifications involve replacing CUB device
primitives with hipCUB equivalents, adapting warp-level
intrinsics, and resolving header conflicts between CUDA and HIP
include paths. This phase preserves the algorithmic core unchanged
and produces a binary-compatible port that the second and third
phases then make work in the AMD environment.

\emph{Build system adaptation.} The CMake build is restructured to
detect the HIP compiler (\texttt{hipcc}) and to target specific GPU
architectures via their GFX identifiers
(\texttt{gfx906} for MI50, \texttt{gfx90a} for MI210, \texttt{gfx942}
for MI300X, and \texttt{gfx1100} for RX~7900~XT). A build-time
option controls whether the library compiles with 32-thread or
64-thread wavefront assumptions, so that the same source tree
serves the CDNA family (fixed wave64) and RDNA~3 (configurable
wave32 or wave64) without runtime dispatch. The per-architecture
binaries are produced as separate shared libraries and shipped in
distinct Singularity images, one per \texttt{gfx<arch>}, so that the
exact target ISA is fixed at container-build time.

\emph{Correctness validation.} The port is validated through a
round-trip check that requires
\texttt{decompress(compress(data))} to be bit-exact equal to the
input across all chunk sizes, data types, and algorithms. This is
the standard correctness test for lossless compression
implementations and ensures that no data corruption is introduced
during the porting process.

\subsection{Algorithm portfolio}
\label{sec:arcto-portfolio}

\arcto{} ships four compression algorithms organized along two
orthogonal axes: input domain (byte-level versus floating-point) and
reconstruction fidelity (lossless versus lossy). The byte-level
lossless tier comprises LZ4, Snappy, and Cascaded, all three
inherited from the \nvcomp{} codebase and hipified through the
process of Section~\ref{sec:arcto-porting}. The three algorithms
cover complementary operating points within the byte-level family:
LZ4 and Snappy prioritize raw throughput with dictionary-style match
finding, while Cascaded layers run-length, delta and bit-packing
stages for inputs whose entropy is concentrated in slow trends rather
than in repeated matches. This tier is the direct hipified equivalent
of the corresponding \nvcomp{} batched offering and exposes the same
scalar-byte input contract.

The floating-point tier is covered by ZFP, integrated into \arcto{}
through a different path than the byte-level codecs. Rather than
hipifying the \texttt{zfp} reference implementation,
\arcto{} carries it as a git submodule so that the algorithm core
remains untouched and trackable to its upstream maintainer; only the
AMD back-end glue and a new GPU-native lossless module are written
within \arcto{}, which ensures that any future change in the upstream
ZFP release is a submodule update rather than a divergent fork.

ZFP is exposed through the four operating modes of its canonical
implementation~\cite{lindstrom2014zfp}, which differ in what they
guarantee about the reconstruction. The \emph{fixed-accuracy} mode
bounds the absolute reconstruction error of every output value by a
user-specified tolerance $\tau$; the \emph{fixed-rate} mode allocates
a fixed bit budget per value and therefore yields a deterministic
output size and a deterministic compression ratio; the
\emph{fixed-precision} mode preserves a user-specified number of
bit planes of the transform and trades them against the tail bits.
The \emph{reversible} mode is the lossless counterpart of the three
lossy modes and is exposed by the API for workflows that mandate
bit-exact reconstruction; it is not exercised in the measurement
campaign of Section~\ref{sec:evaluation}, where the three lossy modes
carry the ZFP story.

The lossy modes are a deliberate offering rather than a fallback. For
seismic forward modelling and the reverse-time-migration workflow it
feeds, bounded-error reconstruction of the wavefield does not
propagate visibly into the migrated image at the tolerances that yield
the largest storage savings, a property established by Lindstrom,
Chen and Lee on full-3D waveform tomography
data~\cite{lindstrom2016f3dt}. Fixed-accuracy is the right operating
point when the error budget is the constraint; fixed-rate is the
right operating point when the storage budget is the constraint;
fixed-precision is reported alongside as a third operating point that
gives explicit control over how many bit planes of the transform are
retained.

The two optimizations of Sections~\ref{sec:optim-chunk-size}
and~\ref{sec:optim-adaptive-tiled} are layered on top of the chunked
batched template that LZ4, Snappy, and Cascaded share, and the
evaluation of Section~\ref{sec:evaluation} reports their effect on
that family. ZFP follows a different execution model: its
block-floating-point transform operates on whole-field three-dimensional
input rather than on the per-chunk windows that the byte-level codecs
consume, so it is benchmarked through a dedicated single-call entry
point and reported on its own axes. The three lossy ZFP modes are
reported on a throughput-versus-quality plane, since their value
proposition combines compression ratio and bounded reconstruction
error rather than throughput alone.

\subsection{Public API and integration model}
\label{sec:arcto-api}

\arcto{} exposes a stable C API that mirrors the nvCOMP batched
signature. The two design points behind the C-not-C++ decision are
ABI stability across compiler versions, which a public C surface
guarantees and a templated C++ interface does not, and direct
linkability from Fortran call sites in production scientific codes.
A typical call site that compresses a batch with LZ4 differs from
its nvCOMP counterpart only in the prefix of the entry-point names
(\texttt{arctoBatchedLZ4*} versus
\texttt{nvcompBatchedLZ4*}); the argument order, types, and
asynchronous semantics are identical, so a CUDA-targeted
application can switch backend with an \texttt{\#ifdef} swap of the
function names. This compatibility is what enables the end-to-end
comparison reported in Section~\ref{sec:evaluation} to be measured
at matched algorithmic configurations.

\subsection{Optimization roadmap}
\label{sec:arcto-optim-roadmap}

The substrate just described is what \arcto{} inherits and extends
on top of the \nvcomp{} port. The two profile-driven optimizations
that constitute the contributions of the present work are layered on
top of this substrate without modifying the algorithmic kernels or
the public ABI. The first, presented in
Section~\ref{sec:optim-chunk-size}, is kernel-side: it replaces the
inherited default chunk size with a per-architecture wave-saturation
budget that fills the CU array in a single launch, and is exposed
through a single launch parameter (\texttt{chunk\_size}) on every
batched entry point. The second, presented in
Section~\ref{sec:optim-adaptive-tiled}, is transfer-side: it
replaces the textbook single-shot pinned-host staging with a
window-based pipeline whose window size is selected by a cost model
and refined online during the call, and is exposed through a new
public entry point (\texttt{arctoHostBatchAdaptive}) that coexists
with the existing single-shot \texttt{arctoHostBatch} variant. Both
changes are additive at the API level and preserve the
binary-compatibility guarantee of
Section~\ref{sec:arcto-constraints}.

\subsection{Repository structure and reproducibility}
\label{sec:arcto-repro}

The source tree splits cleanly into the public headers
(\texttt{include/arcto/}), the algorithm implementations
(\texttt{src/lowlevel/}), the benchmark suite
(\texttt{benchmarks/}), and a test suite driven by \texttt{ctest}
that automates the round-trip correctness checks. The benchmark
binaries are the same artefacts used to produce every CSV reported
in this paper, with hash-stable outputs across architectures for the
lossless cases so that regression testing is meaningful.

The build and execution environment is containerized using
Singularity, the standard container runtime in HPC due to its
rootless execution model and native GPU passthrough. The container
image is based on Ubuntu~22.04 with ROCm~7.0.1 and accepts the GPU
architecture as a build-time argument, so the same software stack is
used across all platforms while still producing
architecture-specific binaries (one Singularity image per
\texttt{gfx<arch>}). Each measurement campaign archives the SIF
image hash alongside the raw CSVs, which combined with the
benchmark-binary commit hash makes the data-plus-code-plus-container
triple that produced any reported figure fully recoverable.
