\section{Optimization~1: Wave-Saturation Chunk Sizing}
\label{sec:optim-chunk-size}

The first \arcto{} optimization is kernel-side: it concerns the chunk
size at which the batched compression and decompression kernels are
dispatched. The decision is purely a launch parameter, hidden from the
algorithmic kernels themselves, and it applies identically to LZ4,
Snappy, and Cascaded. The goal of this section is to make the
chunk-size choice fall out of a per-architecture constant that
\arcto{} can compute once at context-creation time, replacing the
NVIDIA-tuned default inherited from the \nvcomp{} codebase.

\subsection{Why the inherited default under-utilizes AMD GPUs}
\label{sec:chunk-motivation}

The \nvcomp{} batched API was tuned on NVIDIA GPUs whose wavefronts
are 32 threads wide and whose SM counts and L1 sizes differ from their
AMD counterparts. Its default chunk size, on the order of tens of
kibibytes, was selected so that each thread block fits comfortably in
shared memory on Volta-class hardware and so that a batch of a few
thousand chunks is enough to fill the SM array. On AMD wavefronts of
32 lanes (RDNA~3) or 64 lanes (the CDNA family), the same default
produces a chunk count that is several times larger than what the CU
array can absorb in a single wave, and each individual chunk is small
enough that its kernel-startup overhead dominates its useful work.

The characterization study that preceded the present work measured
this effect across the same four architectures targeted
here~\cite{anon_iccsa2026} and found that, at the inherited default,
the AMD kernels achieve roughly half of their attainable throughput on
inputs above a few hundred megabytes. The lost throughput is not
algorithmic; it is structural, and a different launch decision
recovers it without touching the kernels.

\subsection{Wave-saturation as the right-sizing principle}
\label{sec:chunk-formula}

The principle that drives \arcto{}'s chunk-size choice is wave
saturation. A given AMD GPU exposes $N_{\text{CU}}$ compute units,
each able to host $W_{\text{max}}$ resident wavefronts of $w$ threads.
The \emph{wave-slot capacity} of the device is the product

\[
S \;=\; N_{\text{CU}} \cdot W_{\text{max}} \cdot w,
\]

which is the largest number of threads that can be in flight
simultaneously. Once the input is partitioned into a number of chunks
sufficient to occupy $S$ wave slots, additional fragmentation buys no
extra parallelism: it only adds per-chunk metadata, per-launch
overhead, and host-side bookkeeping. Conversely, if the partitioning
produces fewer than $S$ wavefronts, the CU array is left partially
idle for the duration of the launch.

The operational rule, given an input of length $L$ and a target wave
budget $S$, is to pick a chunk size $C$ such that the number of chunks
$\lceil L / C \rceil$ is just enough to fill $S$ once. In practice
\arcto{} captures this through a per-architecture constant
$W_{\text{sat}}$ that absorbs both the wave-slot capacity and the
per-byte wavefront yield of the kernels:

\[
C_{\text{opt}}(L) \;=\;
\max\!\left(W_{\text{sat}},\;
            \left\lceil \frac{L}{\lceil L / W_{\text{sat}}\rceil}\right\rceil
     \right).
\]

The first argument of the $\max$ enforces the per-arch lower bound:
any chunk smaller than $W_{\text{sat}}$ leaves the CU array
under-fed, regardless of how large $L$ is. The second argument is the
input-driven adjustment that keeps the per-launch chunk size
proportional to the input length once it is large enough to matter,
which avoids producing a single oversized chunk on inputs that exceed
$W_{\text{sat}}$ by orders of magnitude.

\subsection{Per-architecture calibration of $W_{\text{sat}}$}
\label{sec:chunk-calibration}

The value of $W_{\text{sat}}$ is determined empirically, once per
architecture, by sweeping the chunk size at fixed input size and
recording the kernel-only throughput. The sweep is bounded below by
$4$\,KiB (the practical minimum for a meaningful kernel launch) and
above by $1$\,GiB (beyond which the kernel becomes
memory-bandwidth-bound and chunk size ceases to matter). The knee of
the throughput-versus-chunk-size curve, defined as the smallest chunk
size that reaches within 5\,\% of the curve's plateau, gives
$W_{\text{sat}}$. Table~\ref{tab:wsat} reports the calibrated values
for the four target architectures.

\begin{table}[t]
\centering
\caption{Calibrated wave-saturation budget per architecture.}
\label{tab:wsat}
\begin{tabular}{lrr}
\toprule
Architecture & CUs & $W_{\text{sat}}$ (MiB) \\
\midrule
gfx906  (MI50)        &  60 & 15 \\
gfx90a  (MI210)       & 104 & 52 \\
gfx942  (MI300X)      & 304 & 76 \\
gfx1100 (RX 7900 XT)  &  84 & 48 \\
\bottomrule
\end{tabular}
\end{table}

The progression from gfx906 through gfx942 tracks the roughly
fivefold growth in CU count across the CDNA generations. The RDNA~3
value sits between gfx90a and gfx942 because the per-CU wavefront
capacity on RDNA~3 is higher (the architecture exposes configurable
wave32 and wave64 modes, and the present measurements use wave32) even
though the absolute CU count is smaller than gfx942's. The single
table is the only piece of \arcto{} that needs to be revisited when a
new AMD generation is targeted; the surrounding formula and the API
exposure are unchanged.

\subsection{Exposure in the \arcto{} API}
\label{sec:chunk-api}

The chunk size is a single launch parameter (\texttt{chunk\_size}) on
every batched entry point in the \arcto{} public API. The library
applies $C_{\text{opt}}$ as the default whenever the application
passes the sentinel value \texttt{0}, and forwards an explicit
non-zero value unchanged. Applications that have an external reason
to pin the chunk size, such as matching a host-side buffer layout or
reproducing a measurement campaign, keep full control. Applications
that do not care receive the wave-saturated value without writing any
code beyond the existing nvCOMP-style call.

The lookup of $W_{\text{sat}}$ from the per-architecture table happens
once at \texttt{arctoCreateContext} time, by querying
\texttt{hipGetDeviceProperties} for the active device's
\texttt{gcnArchName} string. There is no per-launch dispatch on
architecture inside the hot kernels themselves, which preserves the
constraint stated in Section~\ref{sec:arcto-constraints} that
per-architecture tuning must not add branches in the critical path.

\subsection{Measured impact}
\label{sec:chunk-impact}

A preliminary sweep across the four target architectures, whose full
matrix is reported in Section~\ref{sec:evaluation}, shows that
replacing the inherited default with $C_{\text{opt}}$ delivers a
kernel-only speedup of up to $2.87\times$ on gfx942 at the
$16$\,GiB input size, with monotone improvements at smaller inputs
once $L$ exceeds $2 \cdot W_{\text{sat}}$. Below that threshold the
input is too small to fill the CU array regardless of chunk size, and
the formula correctly recovers a single-launch decision (the second
argument of the $\max$ collapses to $L$ itself). The benefit is
algorithm-independent: LZ4, Snappy, and Cascaded all see speedups
within a factor of $1.3\times$ of one another at the same $(L,
\text{arch})$ pair, which confirms the modelling assumption that the
per-byte wavefront yield is roughly constant across the three
byte-level codecs and justifies a single per-arch table rather than a
per-algorithm one.

This first optimization addresses on-GPU compute time only. Even with
a wave-saturated chunk size, the end-to-end pipeline remains dominated
by host-to-device transfer and host-side allocation whenever the input
does not already reside in pinned device-visible memory, which is the
default situation for production scientific codes. The second
optimization, presented in Section~\ref{sec:optim-adaptive-tiled},
addresses that transfer-side cost using a complementary cost model
that tiles the transfer rather than the kernel.
