\section{Optimization~2: Adaptive Tiled Aggregation}
\label{sec:optim-adaptive-tiled}

The second \arcto{} optimization is transfer-side. It targets the
host-to-device pipeline that surrounds the compression kernel and that
dominates end-to-end time once the kernel itself is wave-saturated by
the mechanism of Section~\ref{sec:optim-chunk-size}. The optimization
takes the form of a new public entry point,
\texttt{arctoHostBatchAdaptive}, that coexists with the existing
single-shot \texttt{arctoHostBatch} variant and exposes a window-based
pipeline driven by a cost model and refined online.

\subsection{The single-shot pinned-host trap}
\label{sec:tile-motivation}

The textbook prescription for hiding PCIe latency in a host-driven
GPU workload is to allocate a pinned host buffer of the same size as
the input, memcpy the input into it, transfer the whole buffer to the
device, and launch the kernel. This is what the
\texttt{arctoHostBatch} entry point does, mirroring the equivalent
\nvcomp{} call. Reported in isolation, with the allocation cost
amortized away or treated as a one-off setup expense, this single-shot
pinned variant looks faster than the pageable-source baseline that
lets HIP page through automatically.

Honestly accounted, however, the single-shot variant is a trap. The
pinned-host allocation throughput on the four target platforms is
roughly $220$\,MB/s, two orders of magnitude below the PCIe transfer
bandwidth itself. Allocating a $16$\,GiB pinned buffer therefore
takes about $72$\,seconds before the first byte moves to the device,
which is more than the whole pageable-source pipeline would have
needed. When the allocation cost is included in the end-to-end time,
the single-shot pinned variant is between $0.39\times$ and
$0.59\times$ as fast as the baseline across the four architectures, a
negative speedup that the literature glosses over because the
allocation is treated as outside the measurement window. The
adaptive tiled optimization presented in the remainder of this
section closes this gap and turns the transfer-side hand into a net
positive, but the starting point has to be acknowledged: the apparent
single-shot win is an accounting artefact, not a real one.

\subsection{Cost model for the tiled pipeline}
\label{sec:tile-cost-model}

The adaptive variant replaces the single full-input pinned buffer
with a single fixed-size window of $W$ bytes that is allocated once
and reused for every tile. The input of length $L$ is processed in
$\lceil L/W \rceil$ successive tiles, each of which traverses four
phases:

\begin{enumerate}
  \item a host-to-host memcpy from the pageable source into the
        pinned window, at throughput $R_{\text{h2h}}$
        ($\approx 11$\,GB/s on the target platforms);
  \item a host-to-device PCIe transfer of the pinned window, at
        throughput $R_{\text{pcie}}$ ($\approx 25$\,GB/s);
  \item kernel execution on the tile, of duration $T_k(W)$;
  \item a device-to-host transfer of the compressed output, which is
        small enough compared to the input that it is ignored in the
        model and absorbed by the measurement.
\end{enumerate}

The pinned-window allocation itself, at throughput
$R_{\text{alloc}} \approx 220$\,MB/s, is paid once per call rather
than once per tile. The end-to-end time is therefore

\[
T(L, W) \;\approx\; \frac{W}{R_{\text{alloc}}}
                  \;+\; \left\lceil \frac{L}{W} \right\rceil
                        \cdot T_{\text{tile}}(W),
\]

with the per-tile time, under three-stage overlap of memcpy, PCIe,
and kernel, given by

\[
T_{\text{tile}}(W) \;=\;
   \max\!\left(\frac{W}{R_{\text{h2h}}},\;
               \frac{W}{R_{\text{pcie}}},\;
               T_k(W)\right).
\]

Two constraints on $W$ follow from this structure. The window must be
large enough that the kernel does not stall the pipeline, which gives
$W \geq W_{\text{kernel-sat}}$ with the same per-arch constant as in
Section~\ref{sec:chunk-calibration}. The window must also be large
enough that the PCIe transfer is not dominated by per-transfer setup
overhead, which gives $W \geq W_{\text{pcie-amort}}$, a constant of
about $4$\,MiB on every target platform. A third constraint
follows from the requirement that the allocation cost not eclipse the
useful work: amortizing $W/R_{\text{alloc}}$ over $L/R_{\text{pcie}}$
useful PCIe time gives $W \leq L \cdot R_{\text{alloc}} /
R_{\text{pcie}}$, which is approximately $L/110$. Putting these
together, the operational window choice is

\[
W_{\text{opt}}(L) \;=\;
   \min\!\Big(\frac{L}{110},\;
              \max\!\big(W_{\text{kernel-sat}},\,
                         W_{\text{pcie-amort}}\big)\Big),
\]

where the upper cap from the amortization condition becomes active
only on inputs that are too small to accommodate a fully wave-saturated
window. On the $1$\,GiB to $16$\,GiB inputs that drive the
present study, the lower bound from the kernel-saturation budget is
always the binding constraint.

\subsection{Online refinement}
\label{sec:tile-refinement}

The cost-model constants used to seed $W_{\text{opt}}$ are device
properties (allocation throughput, PCIe bandwidth, kernel saturation
budget) that are stable within a given host-device pair but vary by
roughly $\pm 15\,\%$ across runs because of host load, PCIe topology,
and thermal throttling. \arcto{} therefore refines the cost-model
constants online during the tiled execution itself, using a single
exponential moving average per phase:

\[
\hat{R}_{\phi}^{(t+1)} \;=\;
   (1 - \alpha)\,\hat{R}_{\phi}^{(t)}
   \;+\; \alpha \cdot \frac{W^{(t)}}{T_{\phi}^{(t)}},
\]

with $\phi \in \{\text{h2h},\text{pcie},\text{kernel}\}$ and a smoothing
factor $\alpha = 0.3$ that gives the EMA a half-life of roughly two
tiles. After every fourth tile, $W$ is recomputed from the updated
constants and capped to remain within a factor of $\sqrt{2}$ of its
previous value, which prevents the loop from oscillating on a single
noisy measurement. The first four tiles use the model-seeded $W$
unchanged, so the refinement loop never starves the pipeline of work
to measure.

The full refinement loop adds at most a few microseconds per tile and
is performed on the host side outside the GPU critical path. It is
the only piece of the optimization that depends on the workload
characteristics; the cost-model seeds are static per-arch constants.

\subsection{Exposure in the \arcto{} API}
\label{sec:tile-api}

The adaptive variant lives behind a new public entry point that
mirrors the existing batched signature with two additional
parameters: an opaque adaptive-context handle and an optional
statistics-output pointer. A typical call sequence is:

\begin{lstlisting}[basicstyle=\ttfamily\small,frame=single,xleftmargin=1em]
arctoAdaptiveContext_t ctx;
arctoHostBatchAdaptiveCreate(&ctx, device_id);

arctoHostBatchAdaptiveLZ4(
   ctx, host_input, input_size, ...);

arctoAdaptiveStats_t stats;
arctoHostBatchAdaptiveGetStats(ctx, &stats);
arctoHostBatchAdaptiveDestroy(ctx);
\end{lstlisting}

The context object owns the pinned window, the EMA state, and the
per-arch cost-model seeds. The same context can be reused across
many calls and across different algorithms (LZ4, Snappy, Cascaded),
which lets a long-running scientific simulation amortize the pinned
allocation across the whole simulation rather than per snapshot.
The single-shot \texttt{arctoHostBatch} variant is preserved for
applications that need the simpler signature or that already manage
their own pinned staging area.

The internal implementation lives in
\texttt{src/lowlevel/HostBatchAdaptive.cpp} and is approximately
$240$ source lines, of which the cost-model evaluation and the EMA
refinement loop together account for fewer than $60$ lines. No
changes to the algorithmic kernels are required; the kernels see the
same per-tile chunked workload they would see under any other
window-sizing scheme.

\subsection{Measured impact}
\label{sec:tile-impact}

The adaptive tiled variant gives an end-to-end speedup of $1.32\times$
to $2.26\times$ over the pageable-source baseline on gfx942 across
the $1$\,GiB to $16$\,GiB sweep. The largest observed gain,
$2.26\times$ on the $16$\,GiB Cascaded workload, comes from the
combination of a fully wave-saturated kernel (Section~\ref{sec:optim-chunk-size}),
a single small pinned allocation that the EMA refinement quickly
right-sizes around $80$\,MiB, and three-stage overlap of memcpy,
PCIe, and kernel. The behaviour generalizes qualitatively across the
four architectures: every platform sees a positive speedup at every
input size at or above $1$\,GiB. The two optimizations compose
additively rather than multiplicatively, which is the expected
behaviour given that Section~\ref{sec:optim-chunk-size} reclaims
kernel time and the present section reclaims transfer-pipeline
time. Section~\ref{sec:evaluation} reports the full matrix and breaks
down the contribution of each phase across the four target
architectures.
